EEG Waveform Analysis with applications to Brain Computer Interfaces

### 2018-08-14 sensors.tex
Added math formulae

Overall, EEG signals can be described by their phase, amplitude,  frequency and \textit{waveform}.  The following components regularly  characterize EEG signals:

@@

\begin{itemize}
\item Monomorphic: Distinct EEG activity appearing to be composed of one dominant activity
\item Polymorphic: distinct EEG activity composed of multiple frequencies that combine to form a complex waveform.
\item Transient. An isolated wave or pattern that is distinctly different from background activity.
\end{itemize}

@@

\begin{table}[H]
\caption{Subjective EEG characterization methods found in the surveyed literature.}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{ccc}
\toprule
\textbf{Method}	& \textbf{Phenomena} & \textbf{Reference}	\\
\midrule
SVM     & m-c  & $50\%$ \\
NN       & m-c  & $45\%$ \\
HGO  & PO8 & $25\%$ \\
MP 1 & PO8 & $25\%$ \\
PE     & Cz & $20\%$ \\
RS     & Fz & $15\%$ \\
RS     & m-c  & $15\%$ \\
MP 2 & PO7 & $10\%$ \\
SHCC & P4 & $10\%$ \\
NN s-c & Fz & $10\%$ \\
\bottomrule
\end{tabular}
\label{tab:results}
\end{table}

@@

Other works on EEG explored the idea to extend human capacities analyzing EEG waveforms \citep{Klein1976} where a feature from the amplitude and frequency of its signal and its derivative in time-domain is used.  Moreover, other works explored the use of Mathematical Morphology \citep{Yamaguchi2009}, where the time-domain structure of contractions and dilations were studied. Finally the proposals of Fujimori, Uchida and the PAA \citep{Uchida1996} algorithm are few of the earliest proposals where the idea of capturing the shape of the signal were established.

@@

\textit{Pursuit} algorithms refer, in their many variants, as blind source separation \citep{Comon2010} techniques that assume that the EEG signal is a linear combination of different sparse sources extracted from a template's dictionaries.  \textit{MP}, the most representative of this algorithms, is a greedy variant that decomposes a signal into a linear combination of waveforms, called atoms, that are well localized in time and frequency \citep{ChandranKS2016}.  Given a signal, this optimization technique, tries to find the indexes of atoms and their weights (contributions) that minimize,

@@

which is the error from the signal approximation constructed by the weighted atoms.  The algorithm goes by first setting the approximation signal by the original signal itself,

@@

Hence, it finds the best template, out of the dictionary, that matches current approximation (original signal).

@@

where $g_{yi}$ are all the available scaled, translated and modulated atoms from the dictionary.  This determines the atom selection process, and their contribution is calculating based on

@@

with $k$ representing the index of the selected atom $g_{yk}$.  Finally the contribution of each atom is subtracted from the next approximation, refining the approximation process \citep{Cohen2008,Sanei1997, Mallat1993}

@@

Two variants of this algorithm are evaluated. In \textit{MP 1} the dictionary is directly constructed with the normalized templates from Fig \ref{fig:dictionaryfig}.  In \textit{MP 2} the coefficients of  Daubechies least-asymetric wavelet with 2 vanishing moments atoms are used to construct the dictionary.  The first version, the matching against the template is evaluated according to Equation \ref{eq:mperror} directly, whereas for the latter to craft each feature is crafted by decomposing the signal in its coefficients and building a vector with them.

\begin{equation}
f =  {\bigg ( w_{i} \bigg )}_{1}^{m} 
\label{eq:multiclassificationrow}
\end{equation}

@@

Bond and Pompe Permutation Entropy has been extensively used in EEG processing, particularly for Epilepsy pre-ictal detection \citep{Bandt2002}.  This method generates a code based on the orderly arrangement of sequential samples, and then derives a metric which is based on the amount of entropy of each code within the signal.   For instance, a pure random signal, will achieve the maximum entropy value due to the probability of each code being equal for all of them. This method is the best representative of Waveform Complexity: Lempel-Ziv method L-Z complexity are other variants which use a different definition (Permutation Entropy a new feature for brain computer interfaces).

A signal represented by

\begin{equation}
(x_1,x_2,...,x_{N})
\label{eq:pesignal}
\end{equation}

is resampled by $\tau$ values doing

\begin{equation}
(x_n,x_{n+\tau},x_{n+2 \tau}...,x_{n+(N-1)\tau})
\label{eq:pesignal}
\end{equation}

@@

This method is an extension of Slope Chain Code (SCC) (ref a SCC paper y a SHCC paper), and a code of a sequence of sample points is generated based on the angle between the horizontal segment and any segment, one by one.  This method can be encompass as a syntactic analysis technique, where the EEG segment is represented as a series of elementary patterns, similar to tokens, due to the quantization of angles.

@@

\noindent with $n$ being the original sampling index, and $ 1 \leq t \leq N $.  On the other hand, $\Delta$ can be obtained by

@@

\noindent with $ G < N $. The value $m$ can be derived from

@@

Next step is the normalization of each paired-values according to

@@

\noindent effectively yielding $x,y \in [0,1]$.

@@

This Histogram of Gradient Orientations is based on using Computer Vision techniques to extract directly information from a signal based on the plot of a signal on a 2D image.  This mimicks exactly what a person is doing by visually inspecting the plot, and the waveform.  To do that, the region of the image where the signal is located is divided in a $4 x 4$ block and the signal bidimensional gradients is calculated.  For each block within the patch, a histogram of gradient orientations, 8 circular orientations, is calculated.  This histogram is concatenated for all the 16 blocks and a feature is thus formed.  The details of the method can be found here \citep{Ramele2016}.  This method was implemented using the VLFeat  \citep{Vedaldi2010} Computer Vision libraries.

@@

>>>>

Overall, EEG signals can be described by their phase, amplitude,  frequency and \textit{waveform}.  The following elements regularly  characterize EEG signals:

@@

\begin{itemize}
\item Monomorphic: Distinct EEG activity appearing to be composed of one dominant activity
\item Polymorphic: distinct EEG activity composed of multiple frequencies that combine to form a complex waveform.
\item Transient. An isolated wave or pattern that is distinctly different from background activity.  Define component, source and sensor space.
\end{itemize}

@@

\begin{table}[H]
\caption{Subjective EEG characterization methods found in the surveyed literature.}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{ccc}
\toprule
\textbf{Method}	& \textbf{Phenomena} & \textbf{Reference}	\\
\midrule
Characteristic Shape           & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
Rising and Falling Phase      & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
Pronounced plateau      & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
Ripples and Wiggles     & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
Sinusoidal Shape        & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
Sawtooth                     & Motor Imagery, Sleep &  \citep{Cacioppo2007, Mak2010} \\
Sharpness or Spike-like     & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
Rectangular     & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
Wicket Shape     & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
Peak and Trough Sharpness Ratio     & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
Symmetry between rise and decay phase     & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
Slope Ratio    & Epilepsy &  \citep{Cacioppo2007, Mak2010} \\
Positive vs Negative Ratio    & Sleep K-Complex &  \citep{Cacioppo2007, Mak2010} \\
Base-to-Peak Amplitude     & ERP  &  \citep{Cacioppo2007, Mak2010} \\
Peak-to-Peak Amplitude     & ERP  &  \citep{Cacioppo2007, Mak2010} \\
Latency                                 & ERP  & \citep{Cacioppo2007, Mak2010}  \\
Integrated Activity               & ERP, Epilepsy & \citep{Cacioppo2007, Shah2015} \\
Cross-Correlation                & ERP, Epilepsy, Sleep & \citep{Cacioppo2007, Shah2015} \\
Coupling \\ Cross Frequency,  Phase-Amplitude, Phase-Phase     & Sleep & $20\%$ \\
Period Amplitude Analysis  & ERP, Epilepsy & \citep{Cacioppo2007, Shah2015} \\
\bottomrule
\end{tabular}
\label{tab:results}
\end{table}

@@

Other works on EEG explored the idea to extend human capacities analyzing EEG waveforms \citep{Klein1976} where a feature from the amplitude and frequency of its signal and its derivative in time-domain is used.  Moreover, other works explored the use of Mathematical Morphology \citep{Yamaguchi2009}, where the time-domain structure of contractions and dilations were studied. Finally the proposals of Burch, Fujimori, Uchida and the Period Amplitude Analysis (PAA) \citep{Uchida1996} algorithm are few of the earliest proposals where the idea of capturing the shape of the signal were established.

@@

\textit{Pursuit} algorithms refer, in their many variants, as blind source separation \citep{Vincent2010} techniques that assume that the EEG signal is a linear combination of different sparse sources extracted from a template's dictionaries.  Matching Pursuit \textit{MP}~\citep{Mallat1993}, the most representative of this algorithms, is a greedy variant that decomposes a signal into a linear combination of waveforms, called atoms, that are well localized in time and frequency \citep{ChandranKS2016}.  Given a signal, this optimization technique, tries to find the indexes of atoms and their weights (contributions) that minimize,

@@

which is the error between the signal and their approximation constructed by the weighted $w_i$ atoms $g_{yi}$.  The algorithm goes by first setting the approximation signal by the original signal itself,

@@

Hence, it finds the best template, out of the dictionary, that matches current approximation.

@@

where $g_{yi}$ are all the available scaled, translated and modulated atoms from the dictionary.  This determines the atom selection process, and their contribution is calculated based on

@@

with $k$ represents the index of the selected atom $g_{yk}$.  Finally the contribution of each atom is subtracted from the next approximation,  extracting the approximating weights and atoms for this iteration \citep{Cohen2014,Sanei2007, Mallat1993}

@@

Two variants of this algorithm are evaluated. In \textit{MP 1} the dictionary is directly constructed with the normalized templates from Fig \ref{fig:dictionaryfig}.  In \textit{MP 2} the coefficients of  Daubechies least-asymetric wavelet with 2 vanishing moments atoms are used to construct the dictionary.  The first version, the matching against the template is evaluated according to Equation \ref{eq:mperror} directly, whereas for the latter each feature is crafted by decomposing the signal in its coefficients and building a vector with them.

\begin{equation}
f =  {\bigg ( w_{i} \bigg )}_{1}^{D} 
\label{eq:multiclassificationrow}
\end{equation}

\noindent where $D$ is the size of the dictionary.

@@

Bond and Pompe Permutation Entropy has been extensively used in EEG processing, with applications on Anesthesia, Sleep Stage evaluation and particularly for Epilepsy pre-ictal detection \citep{Bandt2002}.  This method generates a code based on the orderly arrangement of sequential samples, and then derives a metric which is based on the amount of entropy of each code within the signal.   For instance, a pure random signal, will achieve the maximum entropy value due to the probability of each code being equal for all of them~\citep{Nicolaou2010}.

A signal on a window of length $W$ is represented by

\begin{equation}
(x_1,x_2,...,x_{W})
\label{eq:pesignal}
\end{equation}

and it can be resampled by $\tau$ intervals doing

\begin{equation}
(x_n,x_{n+\tau},x_{n+2 \tau}...,x_{n+(m-1)\tau})
\label{eq:pesignal}
\end{equation}

This sequence of order $m$ corresponds to one ordinal pattern $\pi_{n}$, for $n$ varying between $1$ and $W$.  Within the signal window there can be at most $k$ different ordinal (and overlapping) patterns,

\begin{equation}
(\pi_{1},\pi_{2},...,\pi_{k})
\label{eq:pesignal}
\end{equation}

\noindent with $k = W-(m-1) \tau$.  The probability density function for all the available permutations of order $m$ should be $ p = (p_1,p_2,...,p_{m!}) $ with $ \left\lVert p \right\lVert = 1 $.

Hence, the set of time samples map to a new set of $k$ ordinal patterns, and the \textit{pdf} can be estimated by the empirical permutation entropy,

\begin{equation}
P_i = \frac{1}{k} \sum_{s=1}^{k} \left[ \pi_{s} = \pi_{i} \right]
\label{eq:pesignal}
\end{equation}

\noindent with $1 \leq i \leq m!$.  By using the Iverson Bracket, for each $i$ only those ordinal patterns that were effectively found are counted to estimate $p_i$.  The empirical permutation entropy can be estimated from this histogram as,

\begin{equation}
H(p) = \sum_{i=1}^{m!} p_{i} \; log \frac{1}{p_{i}}
\label{eq:pesignal}
\end{equation}

The implemented code was derived from~\citep{Unakafova2013}, and the model description from \citep{Berger2017}.  This procedure will produce a scalar number for the given signal window of size $W$.  To derive a feature out of it, a sliding window procedure must be implemented.  Thus, the length of the feature is $N - (W + \tau (m - 1)) + 1$.

\begin{equation}
f =  {\bigg ( H(p)_{u} \bigg )}_{1}^{N - (W + \tau (m - 1)) + 1}
\label{eq:multiclassificationrow}
\end{equation}

\noindent with $u$ varying on a sample by sample basis along the signal but excluding the first window.

@@

This method is an extension of Slope Chain Code (SCC)~\citep{Alvarado-Gonzalez2016}. A code is generated from a sequence of sample points based on the angle between the horizontal line on a 2D-plane and any segment produced by two consecutive sample points, regarding them as coordinates on that plane.

@@

\noindent with $n$ being the original sampling index, and $ 1 \leq n \leq N $.  On the other hand, $\Delta$ can be obtained by

@@

\noindent with $ G < N $. Lastly, the value $m$ can be derived from

@@

The next step is the normalization of each ordered-pair according to

@@

\noindent effectively yielding $x,y \in [0,1]$. \\

@@

This Histogram of Gradient Orientations is based on using Computer Vision techniques to extract directly information from a signal based on the plot of a signal on a 2D image.  This mimicks exactly what a person is doing by visually inspecting the plot, and the waveform.  To do that, the region of the image where the signal is located is divided in a $4 x 4$ block and the signal's bidimensional gradients is calculated.  For each block within the patch, a histogram of gradient orientations, 8 circular orientations, is calculated.  This histogram is concatenated for all the 16 blocks and a feature is thus formed.  The details of the method can be found here \citep{Ramele2016}.  This method was implemented using the VLFeat  \citep{Vedaldi2010} Computer Vision libraries.

@@

intelligible property
